---
title: 'The faces of #rstats: a (brief) exploration of gender in the R community'
author: "Adnan Fiaz"
output: 
  html_document:
    keep_md: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, message = FALSE)
face_api_key <- "9a103ed6b2bb45ea88f947e4d4f2107c"
```

The representation of women in technology is gaining more and more focus. In the R community this has lead to the foundation of R Ladies, the awarding of a grant by the R Foundation to them and the offering of a scholarship by the useR conference (the latter two are also about inclusion of other minorities). This is a good thing but it does make me wonder: how do we measure success? how will we know these actions have lead to a better situation? And by how much?

This blogpost by Maelle Salmon and this one by Colin Fay inspired me to explore this issue as follows: if we could somehow measure the representation of women currently in our community we can keep track of this. As Mango are organisers of the EARL conference it would also help us assess our own performance.

So I set out to measure the proportion of women among the faces of #rstats. I used Colin's code to first retrieve the profile pictures. 

```{r, collapse=TRUE, message=FALSE}
# Taken from http://colinfay.me/playing-with-rstats-and-microsoft-computer-vision-api/
library(dplyr)
library(rtweet)
token <- create_token( app = "rstats_faces", consumer_key = "zYtn0puxH5zUcCpsOx1TBMKYe", consumer_secret = "WztgDR2MuDxGQ66ulXd6B9sBUvCfYI8OA2PiYAqVBOQrueiuUV")

users <- search_users(q= '#rstats', n = 1000, parse = TRUE) %>%
  unique() %>% 
  mutate(profile_image_url = gsub("_normal", "", profile_image_url))
```

Then I used the Microsoft Face API to tell me if the picture was of a man or a woman. Colin used `lapply` but I thought I'd give `purrr::map` a try.

```{r}
library(httr)
library(purrr)
getGender <- function(myURLs){
  requestURL <- "https://westus.api.cognitive.microsoft.com/face/v1.0/detect"
  parameters <- list(returnFaceAttributes="gender")
  
  result <- map(myURLs, 
                function(myURL) {
                  Sys.sleep(3)
                  POST(requestURL, query=parameters, 
                       add_headers('Content-Type'="application/json", 'Ocp-Apim-Subscription-Key'=face_api_key), 
                       body=sprintf('{"url":"%s"}', myURL)) %>% 
                    content("parsed")
                  } 
                ) %>% 
    # check if a face was detected and no error was returned
    map_chr(~ ifelse(length(.x)>0 && !("error"%in%names(.x)), .x[[1]]$faceAttribute$gender, NA))
  
  return(result)
}

usersWithGender <- users %>% 
  mutate(gender= getGender(profile_image_url)) %>% 
  select(user_id, gender)
```

From Maelle's great collage we already knew not all profile pictures are of people and the Face API does not return any predictions in these cases. There are also some instances the Face API wrongfully didnt detect a face (not even one of our EARL keynote speaker, sorry Hillary Parker). Altogether the Face API detected  `r sum(!is.na(usersWithGender$gender))` faces and the proportion of women among the #rstat faces is `r sprintf("%.01f%%" ,sum(usersWithGender$gender=="female", na.rm=TRUE)/sum(!is.na(usersWithGender$gender))*100)` (click [here](https://s-media-cache-ak0.pinimg.com/736x/57/6d/fe/576dfe57df89ef97cd31fd7601651bc2.jpg) for the pie chart). This is only an estimate and it might not even be a very accurate one but we can compare it with other estimates. If we look at national statistics with respect to gender diversity in the labour force of the technology sector we get the following: [get stats from uk, us, and?]

```{r}

```


[how far are we as a community from the general standard]

Now that we have rudimentary target, let's see how well certain events perform against it. 

```{r}
library(rvest)
earl2016URL<- "https://earlconf.com/2016/london/speakers/"
speakers <- read_html(earl2016URL) %>% 
  html_nodes(".col-md-3 img") %>% 
  html_attr("src")

earl2016Genders <- getGender(paste0(earl2016URL, speakers))
```

```{r}
library(jsonlite)
earlLondon2017 <- jsonlite::read_json("https://earlconf.com/london/londonSpeakers.json")
earlLondon2017Genders <- earlLondon2017 %>% 
  map("headshot") %>% 
  map(~paste0("https://earlconf.com/img/headshots/london/", .x)) %>% 
  map_chr(~getGender(.))

earlSF2017 <- jsonlite::read_json("https://earlconf.com/sanfrancisco/sanFranciscoSpeakers.json")
earlSF2017Genders <- earlSF2017 %>% 
  map("headshot") %>% 
  map(~paste0("https://earlconf.com/img/headshots/sanfrancisco/", .x)) %>% 
  map_chr(~getGender(.))
```

```{r}
user2016URL <- "http://schedule.user2016.org/directory/speakers"
user2016Genders <- read_html(user2016URL) %>% 
  html_nodes("img") %>% 
  html_attr("src") %>% 
  map(~paste0("http:", .x)) %>% 
  map_chr(~getGender(.))
```

```{r}
myData <- data.frame(Type = c("#rstats faces", "EARL London 2016", "UseR 2016", "EARL SF 2017", "EARL London 2017"),
           PercentageWomen= c(0.23, 5/43, 20/84, 10/30, 11/34))
library(ggplot2)
ggplot(myData , aes(x=Type, y=PercentageWomen)) + geom_bar(stat="identity") + ylim(c(0,1))
```